---
title: "Analysis of Trachymyrmex septentrionalis specificiity"
author: K. Beigel
date: `r Sys.Date()`
---

# Set up using `renv`
```{r}
# Set up 
renv::restore()
```

# Install `coevo` package using `devtools::install_github()`
```{r}
devtools::install_github('kbeigel/coevo@v1.0.0', force = TRUE)
library(coevo)
```

```{r}
set.seed(42)
```

# Analysis for cophylogenetic signal (host: ants, symbiont: fungus)
```{r}
analysis_name = 'antHost_fungusSymbiont'
```

## Read in host-symbiont association matrix (rows: ants, columns: fungus)
```{r}
hs_assoc = read_associations('data/ant-fungus_matrix.tsv')
```

## Read in host and symbiont trees, filtering to keep taxa in the association matrix
```{r}
# Ant phylo
ants = read_phylo(
    filepath = 'data/Trachymyrmex_septentrionalis_GBS-SNP_Beigel2025.nexus',
    nexus_format = 'raxml', 
    taxa_to_keep = hs_assoc$host_taxa,
    label_fix = c('JS_', '')
)

# Fungus phylo
fungus = read_phylo(
    filepath = 'data/Leucocoprinus_sp_Sanger-ITS_Luiso2020.nexus',
    nexus_format = 'raxml', 
    taxa_to_keep = hs_assoc$symbiont_taxa,
    label_fix = c('\\.', '')
)
```
___

## Run ParaFit
```{r}
parafit_results = parafit_analysis(
    host_tree = ants,
    symbiont_tree = fungus,
    assoc_mat = hs_assoc,
    n_runs = 100,
    correction_method = 'lingoes'
)
```

## Run PACo
```{r}
paco_results = paco_analysis(
    host_tree = ants,
    symbiont_tree = fungus,
    assoc_mat = hs_assoc,
    n_perm = 100000
)
```

## Compile the results from ParaFit and PACo
```{r}

# Get a table of the global results for ParaFit and PACo
global_results = compile_global_results(
    parafit_results, paco_results
)

# Get a table of the link results
link_results = compile_link_results(
    parafit_results,
    paco_results,
    p_alpha = 0.05
)

# Write results to file
dir.create('results/tables/', showWarnings = FALSE, recursive = FALSE)
write.table(global_results, paste0('results/tables/', analysis_name, '_coevo_global_results.txt'), sep = '\t', quote = FALSE, row.names = FALSE)
write.table(link_results, paste0('results/tables/', analysis_name, '_coevo_link_results.txt'), sep = '\t', quote = FALSE, row.names = FALSE)
```

## Plotting the cophylo
```{r}
# Create cophylogeny and write to file
dir.create('results/figures/', showWarnings = FALSE, recursive = FALSE)

# Plot cophylogeny
plot_cophylo(
    h_phy = ants,
    s_phy = fungus,
    results = link_results,
    filename = paste0('results/figures/', analysis_name, '_coevo_cophylo'),
    plot_h = 10, 
    plot_w = 14,
    palette = c('gold', 'magenta3', 'cyan3')
)
```

___

# Analysis for cophylogenetic signal (host: fungus, symbiont: ants)

```{r}
analysis_name = 'fungusHost_antSymbiont'
```

## Read in host-symbiont association matrix (rows: fungus, columns: ants)
```{r}
hs_assoc = read_associations('data/fungus-ant_matrix.tsv')
```

## Read in host and symbiont trees, filtering to keep taxa in the association matrix
```{r}
# Ant phylo
ants = read_phylo(
    filepath = 'data/Trachymyrmex_septentrionalis_GBS-SNP_Beigel2025.nexus',
    nexus_format = 'raxml', 
    taxa_to_keep = hs_assoc$symbiont_taxa,
    label_fix = c('JS_', '')
)

# Fungus phylo
fungus = read_phylo(
    filepath = 'data/Leucocoprinus_sp_Sanger-ITS_Luiso2020.nexus',
    nexus_format = 'raxml', 
    taxa_to_keep = hs_assoc$host_taxa,
    label_fix = c('\\.', '')
)
```

## Run ParaFit
```{r}
parafit_results = parafit_analysis(
    host_tree = fungus,
    symbiont_tree = ants,
    assoc_mat = hs_assoc,
    n_runs = 100,
    correction_method = 'lingoes'
)
```

## Run PACo
```{r}
paco_results = paco_analysis(
    host_tree = fungus,
    symbiont_tree = ants,
    assoc_mat = hs_assoc,
    n_perm = 100000
)
```

## Compile the results from ParaFit and PACo
```{r}
# Get a table of the global results for ParaFit and PACo
global_results = compile_global_results(
    parafit_results, paco_results
)

# Get a table of the link results
link_results = compile_link_results(
    parafit_results,
    paco_results,
    p_alpha = 0.05
)

# Write results to file
dir.create('results/tables/', showWarnings = FALSE, recursive = FALSE)
write.table(global_results, paste0('results/tables/', analysis_name, '_coevo_global_results.txt'), sep = '\t', quote = FALSE, row.names = FALSE)
write.table(link_results, paste0('results/tables/', analysis_name, '_coevo_link_results.txt'), sep = '\t', quote = FALSE, row.names = FALSE)
```

## Plotting the cophylo
```{r}
# Create cophylogeny and write to file
dir.create('results/figures/', showWarnings = FALSE, recursive = FALSE)

# Plot cophylogeny
plot_cophylo(
    h_phy = fungus,
    s_phy = ants,
    results = link_results,
    filename = paste0('results/figures/', analysis_name, '_coevo_cophylo'),
    plot_h = 10, 
    plot_w = 14,
    palette = c('gold', 'magenta3', 'cyan3')
)
```

___

# Comparing results
```{r}

fh_as_global = read.table(paste0('results/tables/', 'fungusHost_antSymbiont', '_coevo_global_results.txt'), sep = '\t', header = TRUE)
fh_as_links = read.table(paste0('results/tables/', 'fungusHost_antSymbiont', '_coevo_link_results.txt'), sep = '\t', header = TRUE)

ah_fs_global = read.table(paste0('results/tables/', 'antHost_fungusSymbiont', '_coevo_global_results.txt'), sep = '\t', header = TRUE)
ah_fs_links = read.table(paste0('results/tables/', 'antHost_fungusSymbiont', '_coevo_link_results.txt'), sep = '\t', header = TRUE)

# We can also look at the ParaFit "fungusHost_antSymbiont" and PACo "antHost_fungusSymbiont"
# to see what overlaps between these analyses
res_combo = cbind(fh_as_links[, c(1:6, 9:10)], ah_fs_links[, c(7, 8, 11)])
plot_cophylo(
    h_phy = fungus,
    s_phy = ants,
    results = res_combo,
    filename = paste0('results/figures/', 'CombinedRes_PACoAntHost_ParaFitFungusHost', '_coevo_cophylo'),
    plot_h = 10, 
    plot_w = 14,
    palette = c('gold', 'magenta3', 'cyan3')
)

# We can also look at the ParaFit "antHost_antSymbiont" and PACo "fungusHost_antSymbiont"
# to see what overlaps between these analyses
res_combo = cbind(ah_fs_links[, c(1:6, 9:10)], fh_as_links[, c(7, 8, 11)])
plot_cophylo(
    h_phy = fungus,
    s_phy = ants,
    results = res_combo,
    filename = paste0('results/figures/', 'CombinedRes_PACoFungusHost_ParaFitAntHost', '_coevo_cophylo'),
    plot_h = 10, 
    plot_w = 14,
    palette = c('gold', 'magenta3', 'cyan3')
)

```